\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-07-18 analysis/paper/paper.tex
cleaned cs1

\author{
Sahir Bhatnagar *\\McGill University \And Maxime Turgeon *\\University
of Manitoba \AND Jesse Islam\\McGill University \And James
Hanley\\McGill University \And Olli Saarela\\University of Toronto
}
\title{\pkg{casebase}: An Alternative Framework For Survival Analysis}

\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, Jesse Islam, James
Hanley, Olli Saarela}
\Plaintitle{casebase: An Alternative Framework For Survival Analysis}
\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
Analysis}

@@

\Address{
    Sahir Bhatnagar *\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{https://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

@@

\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.\]

@@

\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]

@@

\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]

@@

@@

For the first case study, we will use of the European Randomized Study
of Prostate Cancer Screening data. This dataset is available through the
\pkg{casebase} package:

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
R+   levels = c(0, 1),
R+   labels = c("Control group", "Screening group")
R+ )
\end{CodeInput}
\end{CodeChunk}

The results of this study were published by
\citep{schroder2009screening}, and the dataset itself was obtained using
the approach described in \citep{liu2014recovering}.

Population time plots can be extremely informative graphical displays of
survival data. They should be the first step in an exploratory data
analysis. We facilitate this task in the \pkg{casebase} package using
the \code{popTime} function. We first create the necessary dataset for
producing the population time plots, and we can generate the plot by
using the corresponding \code{plot} method:

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
R> plot(pt_object)
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}

@@

We can also create exposure stratified plots by specifying the
\code{exposure} argument in the \code{popTime} function:

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC,
R+   event = "DeadOfPrCa",
R+   exposure = "ScrArm"
R+ )
R> plot(pt_object_strat)
\end{CodeInput}

We can also plot them side-by-side using the \code{ncol} argument:

\begin{CodeInput}
R> plot(pt_object_strat, facet.params = list(ncol = 2))
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}

ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.

Next, we investigate the differences between the control and the
screening arms. A common choice for this type of analysis is to use a
Cox regression model and estimate the hazard ratio for the screening
group (relative to the control group). In \proglang{R}, it can be done
as follows:

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
R+   data = ERSPC
R+ )
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

We can also plot the cumulative incidence function (CIF) for each group.
However, this involves estimating the baseline hazard, which Cox
regression treated as a nuisance parameter.

\begin{CodeInput}
R> new_data <- data.frame(
R+   ScrArm = c("Control group", "Screening group"),
R+   ignore = 99
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> legend("topleft",
R+   legend = c("Control group", "Screening group"),
R+   col = c("red", "blue"),
R+   lty = c(1, 1),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}

Next, we show how case-base sampling can be used to carry out a similar
analysis. We fit several models that differ in how we choose to model
time.

First, we will fit an exponential model. Recall that this corresponds to
excluding time from the linear predictor.

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
R+   data = ERSPC,
R+   ratio = 100
R+ )
R> 
R> summary(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.148  -0.148  -0.148  -0.133   3.078  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059  on 54539  degrees of freedom
#> Residual deviance: 6053  on 54538  degrees of freedom
#> AIC: 6057
#> 
#> Number of Fisher Scoring iterations: 7
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
#> ScrArmScreening group 
#>                  0.81
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2, ])
\end{CodeInput}

\begin{CodeOutput}
#>  2.5 % 97.5 % 
#>   0.68   0.96
\end{CodeOutput}
\end{CodeChunk}

We can then use the \code{absoluteRisk} function to get an estimate of
the cumulative incidence curve for a specific covariate profile. In the
plot below, we overlay the estimated CIF from the exponential model on
the Cox model CIF:

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(
R+   object = casebase_exponential,
R+   time = seq(0, 15, 0.1),
R+   newdata = new_data
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft",
R+   legend = c(
R+     "Control group (Cox)", "Control group (Casebase)",
R+     "Screening group (Cox)", "Screening group (Casebase)"
R+   ),
R+   col = c("red", "red", "blue", "blue"),
R+   lty = c(1, 2, 1, 2),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}

As we can see, the exponential model gives us an estimate of the hazard
ratio that is similar to the one obtained using Cox regression. However,
the CIF estimates are quite different. Based on what we observed in the
population time plot, where more events are observed later on in time,
we do not expect a constant hazard would be a good description for this
data. In other words, this poor fit for the exponential hazard is
expected. A constant hazard model would overestimate the cumulative
incidence earlier on in time, and underestimate it later on; this is
what we see on the cumulative incidence plot. This example demonstrates
the benefits of population time plots as an exploratory analysis tool.

For the next model, we include time as a linear term. Recall that this
corresponds to a Weibull model.

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
R+   data = ERSPC,
R+   ratio = 100
R+ )
R> 
R> summary(casebase_time)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.401  -0.162  -0.122  -0.094   3.466  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059  on 54539  degrees of freedom
#> Residual deviance: 5811  on 54537  degrees of freedom
#> AIC: 5817
#> 
#> Number of Fisher Scoring iterations: 8
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
#>               0.00012               1.24882               0.79297
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>                          2.5 %  97.5 %
#> (Intercept)           0.000096 0.00015
#> Follow.Up.Time        1.213876 1.28507
#> ScrArmScreening group 0.665769 0.94256
\end{CodeOutput}
\end{CodeChunk}

Again, the estimate of the hazard ratio is similar to that obtained from
Cox regression. We then look at the estimate of the CIF:

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(
R+   object = casebase_time,
R+   time = seq(0, 15, 0.1),
R+   newdata = new_data
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R+   legend = c(
R+     "Control group (Cox)", "Control group (Casebase)",
R+     "Screening group (Cox)", "Screening group (Casebase)"
R+   ),
R+   col = c("red", "red", "blue", "blue"),
R+   lty = c(1, 2, 1, 2),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}

We see that the Weibull model leads to a better fit of the CIF than the
Exponential model, when compared with the semi-parametric approach.

Finally, we can model the hazard as a smooth function of time using the
\code{splines} package:

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
R+   data = ERSPC,
R+   ratio = 100
R+ )
R> 
R> summary(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.405  -0.173  -0.136  -0.085   3.796  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value             Pr(>|z|)    
#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5790.5  on 54535  degrees of freedom
#> AIC: 5800
#> 
#> Number of Fisher Scoring iterations: 8
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
#>              0.000035             73.578176              8.121029 
#>   bs(Follow.Up.Time)3 ScrArmScreening group 
#>             95.053534              0.805017
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>                           2.5 %     97.5 %
#> (Intercept)            0.000018   0.000063
#> bs(Follow.Up.Time)1   15.835008 367.061126
#> bs(Follow.Up.Time)2    3.252368  21.652009
#> bs(Follow.Up.Time)3   22.537699 364.141700
#> ScrArmScreening group  0.675920   0.956823
\end{CodeOutput}
\end{CodeChunk}

Once again, we can see that the estimate of the hazard ratio is similar
to that from the Cox regression. We then look at the estimate of the
CIF:

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(
R+   object = casebase_splines,
R+   time = seq(0, 15, 0.1),
R+   newdata = new_data
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R+   legend = c(
R+     "Control group (Cox)", "Control group (Casebase)",
R+     "Screening group (Cox)", "Screening group (Casebase)"
R+   ),
R+   col = c("red", "red", "blue", "blue"),
R+   lty = c(1, 2, 1, 2),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}

Qualitatively, the combination of splines and case-base sampling
produces the parametric model that most closely resembles the Cox model.
In the following table, we see that the confidence intervals are also
similar across all four models. In other words, this reinforces the idea
that, under proportional hazards, we do not need to model the full
hazard to obtain reliable estimates of the hazard ratio. Of course,
different parametric models for the hazards give rise to qualitatively
different estimates for the CIF.

\begin{CodeChunk}
\begin{table}[H]
\centering
\begin{tabular}{lrlr}
\toprule
Model & HR & CI & Width\\
\midrule
Cox & 0.80 & (0.67, 0.95) & 0.28\\
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
Weibull & 0.79 & (0.67, 0.94) & 0.28\\
Splines & 0.81 & (0.68, 0.96) & 0.28\\
\bottomrule
\end{tabular}
\end{table}

As noted above, the usual asymptotic results hold for likelihood ratio
tests built using case-base sampling models. Therefore, we can easily
test the null hypothesis that the exponential model is just as good as
the larger (in terms of number of parameters) splines model.

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5790  3      262 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit. Similarly,
we can compare the AIC for all three case-base models:

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R+   "Weib." = AIC(casebase_time),
R+   "Splines" = AIC(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>    Exp.   Weib. Splines 
#>    6057    5817    5800
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Description & Statistical summary\\
\midrule
\code{Sex} & Sex & M=Male (100)\\
 &  & F=Female (77)\\
\code{D} & Disease & ALL (73)\\
 &  & AML (104)\\
\code{Phase} & Phase & CR1 (47)\\
\addlinespace
 &  & CR2 (45)\\
 &  & CR3 (12)\\
 &  & Relapse (73)\\
\code{Source} & Type of transplant & BM+PB (21)\\
 &  & PB (156)\\
\addlinespace
\code{Age} & Age of patient (years) & 4--62\\
 &  & 30.47 (13.04)\\
\code{ftime} & Failure time (months) & 0.13--131.77\\
 &  & 20.28 (30.78)\\
\code{Status} & Status indicator & 0=censored (46)\\
\addlinespace
 &  & 1=relapse (56)\\
 &  & 2=competing event (75)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R+   Status ~ ftime + Sex + D + Phase + Source + Age,
R+   data = bmtcrr,
R+   ratio = 100,
R+   time = "ftime"
R+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+   data = bmtcrr
R+ )
\end{CodeInput}
\end{CodeChunk}

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame(
R+   "Sex" = factor(c("F", "F"),
R+     levels = levels(bmtcrr[, "Sex"])
R+   ),
R+   "D" = c("ALL", "AML"), # Both diseases
R+   "Phase" = factor(c("Relapse", "Relapse"),
R+     levels = levels(bmtcrr[, "Phase"])
R+   ),
R+   "Age" = c(35, 35),
R+   "Source" = factor(c("PB", "PB"),
R+     levels = levels(bmtcrr[, "Source"])
R+   )
R+ )
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(
R+   object = model_cb, time = time_points,
R+   method = "numerical", newdata = newdata
R+ )
\end{CodeInput}
\end{CodeChunk}

>>>>

\author{
Sahir Rai Bhatnagar *\\McGill University \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
}
\title{\pkg{casebase}: An Alternative Framework For Survival Analysis}

\Plainauthor{Sahir Rai Bhatnagar *, Maxime Turgeon *, Jesse Islam, James Hanley, Olli Saarela}
\Plaintitle{casebase: An Alternative Framework For Survival Analysis}
\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}

@@

\Address{
    Sahir Rai Bhatnagar *\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{https://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. \label{eq:gomp}
\end{equation}

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \label{eq:weibull}
\end{equation}

@@

The case-base sampling approach described in Section \ref{theory} can be
visualized in the form of a population time plot. They are extremely
informative graphical displays of survival data and should be one of the
first steps in an exploratory data analysis. We facilitate this task in
the \pkg{casebase} package using the \code{popTime} function and
\code{plot} method as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The \code{casebase::popTime} function takes as input the original
  dataset along with the column names corresponding to the timescale,
  the event status and an exposure group of interest (optional). This
  will create an object of class \code{popTime}.\\
\item
  The corresponding \code{plot} method for the object created in Step 1
  can be called to create the population time plot with several options
  for customizing the aesthetics.
\end{enumerate}

The splitting of these tasks gives the user the flexibility to create
population-time plots with the graphics system of their choice using the
object created in Step 1, while the object created in Step 2 allows one
to leverage the \pkg{ggplot2} \citep{ggplot2} family of functions for
further customization.

To illustrate these functions we will use the European Randomized Study
of Prostate Cancer Screening (ERSPC) data \citep{schroder2009screening}
which was extracted using the approach described in
\citep{liu2014recovering} and available through the \pkg{casebase}
package. This data contains the individual observations for 159,893 men
between the ages of 55 and 69 years at entry who were recruited from
seven European countries.

We first create the necessary dataset for producing the population time
plot using the \code{casebase::popTime} function. In this example we
stratify by the treatment group. The resulting object inherits from
class \texttt{popTime} and stores the exposure variable as an attribute:

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, 
R+                                time = "Follow.Up.Time",
R+                                event = "DeadOfPrCa",
R+                                exposure = "ScrArm")
R> inherits(pt_object, "popTime")
\end{CodeInput}

\begin{CodeOutput}
#> [1] TRUE
\end{CodeOutput}

\begin{CodeInput}
R> attr(pt_object, "exposure")
\end{CodeInput}

\begin{CodeOutput}
#> [1] "ScrArm"
\end{CodeOutput}
\end{CodeChunk}

We then pass this object to the corresponding \code{plot} method which
leverages the \pkg{ggplot2} package to build population time plots.

\begin{CodeInput}
R> plot(pt_object, add.base.series = TRUE)
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a unique \emph{step
shape} in the population time area which is due to the randomization of
the Finnish cohorts which were carried out on January 1 of each of the 4
years 1996 to 1999. This, coupled with the uniform December 31 2006
censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8
years of follow-up. Tracked backwards in time (i.e.~from right to left),
the population-time plot shows the recruitment pattern from its
beginning in 1991, and in particular the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
close their recruitment - in December 2003 - the minimum potential
follow-up is three years. Tracked further forwards in time (i.e.~after
year 3) the attrition is a combination of deaths and staggered entries.

Next, layers for the case series and base series are added. The case
series is sampled at random vertically on the plot to avoid having all
points along the upper edge of the plot, while the base series is
sampled horizontally with sampling weight proportional to their
follow-up time. By randomly distributing the points, we can get a better
sense of the incidence density. In the ERSPC data, we see that more
events are observed later on in time. Therefore a constant hazard model
would not be appropriate in this instance as it would overestimate the
cumulative incidence earlier on in time, and underestimate it later on.
The reader is referred to the package vignettes for more examples and a
detailed description of how to modify the aesthetics of a population
time plot.

@@

We investigate the differences in survival between the control and
screening arms in the ERSPC study presented earlier using casebase
sampling to fit a parametric hazard function. We fit four models that
differ in the functional form of time; (1) excluded from the linear
predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}),
(3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function
using cubic B-splines. The models are fit using \code{fitSmoothHazard}.
Since the output object from \code{fitSmoothHazard} inherits from the
\code{glm} class, we see a familiar result when using the function
\code{summary}:

@@

\begin{CodeInput}
R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
R>              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
R>              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
R>              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
R> 
R> fits <- lapply(fmla, function(i) {
R>   casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
R> })
R> 
R> summary(fits[[2]])
\end{CodeInput}
\end{CodeChunk}

The \code{absoluteRisk} function takes as input the
\code{fitSmoothHazard} object and returns a matrix where each column
corresponds to the covariate profiles specified in the \code{newdata}
argument, and each row corresponds to a specified time point:

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
R> 
R> risks <- lapply(fits, function(i) {
R>   casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
R> })
R> 
R> head(risks[[2]])
\end{CodeInput}
\end{CodeChunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
casebase on the Cox model CIF. The CIF estimates for the exponential
model in panel (1) overestimate the cumulative incidence earlier on in
time, and underestimate it later on. Based on what we observed in the
population time plot, where more events are observed later on in time,
this poor fit for the exponential hazard is expected. This example
demonstrates the benefits of population time plots as an exploratory
analysis tool.

In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit
with increasing complexities of time. As noted above, the usual
asymptotic results hold for likelihood ratio tests built using casebase
sampling models. Therefore, we can easily test the null hypothesis that
the exponential model is just as good as the larger (in terms of number
of parameters) splines model:

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5788  3      265 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit. Similarly,
we can compare the AIC for all four casebase models:

\begin{CodeOutput}
#>    Exp.   Gomp.   Weib. Splines 
#>    6057    5818    6057    5798
\end{CodeOutput}
\end{CodeChunk}

{\centering \includegraphics{../figures/erspc-cox-cif-1}

\caption{Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (2) Linear function of time which is equivalent to a Gompertz distribution. (3) The natural logarithm of time which is equivalent to a Weibull distribution. (4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{CodeChunk}

In the following table, we present a side-by-side comparison of the
hazard ratios and confidence intervals estimated from
\code{fitSmoothHazard} and the corresponding parametric model using
\code{survival::survreg}, as well as the Cox model estimate. The hazard
ratio estimates and confidence intervals are similar across all four
models. This reinforces the idea that, under proportional hazards, we do
not need to model the full hazard to obtain reliable estimates of the
hazard ratio. Of course, different parametric models for the hazards
give rise to qualitatively different estimates for the CIF as shown in
Figure \ref{fig:erspc-cox-cif}.

\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
\centering
\begin{tabular}[t]{>{\bfseries}lcc}
\toprule
Model & casebase::fitSmoothHazard & survival::survreg\\
\midrule
Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
Weibull & 0.81 (0.68, 0.97) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
Splines & 0.80 (0.67, 0.95) & --\\
\bottomrule
\multicolumn{3}{l}{Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
\end{tabular}
\end{table}

@@

@@

@@

@@

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R>   data = bmtcrr
R> )
\end{CodeInput}
\end{CodeChunk}

@@
